### 2018-03-17 paper/3-abstractions.tex
Minor changes
Add a relaxed correctness definition for shallow builds
Proof S3

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in the previous
section~\S\ref{sec-background} and design complex build systems from simple
primitives. Specifically, we present the \emph{task} and \emph{build}
abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We  begin by establishing a common vocabulary for build systems.

\vspace{-2mm}
\paragraph{Keys, values, and the store} The goal of any build system is to
bring up to date a \emph{store} that implements a mapping from \emph{keys} to
\emph{values}. In software build systems the store is the file system, the
keys are filenames, and the values are file contents. In \Excel, the store is
the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
numbers, strings etc, displayed as the cell contents. Many systems use
\emph{hashes} of values as compact summaries with a fast equality check.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
has a task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

@@

\noindent
As one can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
\hs{fetch}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
formula.

@@

Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to
\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
one another.

@@

\noindent
Here, \cmd{B1} and \cmd{B2} statically depend on each other, but \Excel (which
uses dynamic dependencies) is perfectly happy. We can express this using
our task abstraction as follows:

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
\hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies, and hence
monadic tasks, are much more challenging, as we will see
in~\S\ref{sec-implementations}.

@@

    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{Just}~\hs{(}\hs{getValue}~\hs{result}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

@@

\noindent
The implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:

>>>>

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed previously in \S\ref{sec-background},
and design complex build systems from simple
primitives. Specifically, we present the \emph{task} and \emph{build}
abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems:

\vspace{-2mm}
\paragraph{Keys, values, and the store} The goal of any build system is to
bring up to date a \emph{store} that implements a mapping from \emph{keys} to
\emph{values}. In software build systems the store is the file system, the
keys are filenames, and the values are file contents. In \Excel, the store is
the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
numbers, strings etc, displayed as the cell contents. Many build systems use
\emph{hashes} of values as compact summaries with a fast equality check.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
has a task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

@@

\noindent
As we can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
\hs{fetch}~\hs{"B1"} twice -- once per occurrence of \cmd{B1} in the
formula.

@@

The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to
\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
one another.

@@

\noindent
Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
uses dynamic dependencies) is perfectly happy. We can express these rules using
our task abstraction as:

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} \emph{different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies, and hence
monadic tasks, are much more challenging, as we shall see
in~\S\ref{sec-implementations}.

@@

    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:
